\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-11-03 paper.tex

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress, however, has led to a
% Controversial, perhaps, but true?
substantial gap opening between theory and practice.
Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
based on exhaustively detailing the ``events'' that occur
in the history of a sample,
are insufficient to describe the outputs of current methods.
Moreover, we argue that the underlying assumption
that all events can be known and precisely estimated
is fundamentally unsuited to the realities
of modern, population-scale datasets.
We propose an alternative mathematical formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

% What's the problem?
A major problem currently facing the field is that
classical mathematical formalisms and terminology
cannot adequately describe these vast inferred ARGs.
Fundamentally, these formalisms assume that an
ARG is known in complete detail
and are not suited to describing partial or approximate knowledge.
As we are actively inferring ARGs at the population scale,
and such ARGs can never be known in complete detail,
there is currently a substantial gap between our theoretical frameworks
and practical application.
The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% This is a digression from the maing point of the paragraph, so
% making it a parenthetical aside.
(Note that it is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.)
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

@@

% Moving on from ARG-unary to the more general case
More generally, locally unary nodes,
which can have one or more children in the graph,
are a common and important feature
of many different types of ARG.
As discussed in the previous section,
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them. As we see in the next two sections, both ARG and locally
unary nodes occur in various scenarios, and are
produced by a range of current inference methods.

@@

% Here we describe recombination features
Considering Fig.~\ref{fig-inferred-args}, we can see that
there is substantial variation in the number of recombination breakpoints
inferred by different methods, with e.g.~\argweaver\ suggesting
far more than the 7 required for this dataset under minimal parsimony
assumptions~\citep{song2003parsimonious}.
A sense of the amount of recombination in each ARG is provided by
the node colouring scheme, which shows the number of parents for each node.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event
corresponds to a node with exactly two parents and one child.
As these methods explicitly infer a recombination event for each
breakpoint, the number of breakpoints equals the number of two-parent
(brown) nodes.
In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
straightforward
relationship between the number of nodes with multiple parents
and number of breakpoints along the genome.
In both ARGs the number of breakpoints is smaller
than the number of multiple-parent ARG nodes, showing
that several multiple-parent nodes must share breakpoint positions.
There are also ARG nodes with multiple parents and multiple children,
where one or more recombinations have been pushed down onto a
more recent node.
As a consequence, it may be difficult to condense each transition between trees
in these ARGs into a set of SPR operations.

@@

\section{Discussion}
\label{sec-discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in inference methods, their evaluation, and their application to biological questions.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix~\ref{sec-arg-history}, the term
``ancestral recombination graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition
within the context of stochastic
processes and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful; we have
tried to clarify and systematise it here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG) encoding, which is more limited in what can be described.
Importantly, gARGs do not require fully precise estimates of
ancestral recombination events,
and allow us to directly express important forms of temporal uncertainty.

Fully decoupling the general concept of an ARG from the coalescent
with recombination (henceforth, ``coalescent'')
is an important step.
While the coalescent has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
One key assumption is that
sample size $n$ is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,
karczewski2020mutational,tanjo2021practical,
halldorsson2022sequences},
and so sample size is an order of magnitude \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from this
assumption, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
A model assuming a single $N_e$ would be a
drastic over-simplification of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the complexities
of family structure in these
datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
Another core assumption of the coalescent model is that the genome
(or at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times.
High-quality whole genome assemblies are now available
for many species % Can we say something more concrete here?
and projects are under way to obtain
them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
and so we can expect inferred ARGs to routinely span
large fractions of a chromosome.

Recent large-scale methods have simplified the inference problem by
making a single, deterministic best-guess at ARG
inference~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes, the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast.
Thus, although it is clearly an oversimplification to arbitrarily
choose \emph{one} best guess,
it is not clear that generating many guesses
when sample sizes are large will achieve much.
At the scale of millions of samples,
we could only ever explore the tiniest corner of the incomprehensibly
large space of plausible ARGs.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference, and to
incorporate uncertainty encountered during inference into the returned ARG.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have more than two children (polytomies
representing uncertainty over the ordering of coalescence events, Appendix~\ref{sec-cell-lineages-and-args}) or
more than two parents (representing uncertainty over the ordering
of multiple recombination events, Section~\ref{sec-ARG-simplification}). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.
How this uncertainty can be utilised in downstream applications
is an open question.

@@

The schematic illustrates an important point about the biological reality of polytomies.
Three linages coalesce in the left-hand genome of individual $D_{10}$, but do
so as the result of two successive bifurcations. This is \emph{necessarily} so,
because the only known method of reproducing DNA is by (semi-conservative) duplication.
Whether this polytomy is resolvable depends on the available mutational data.
Mutations can occur along any cell lineages. For example,
a mutation in the first cell division of $D_{10}$ could be shared between
the two gametes produced by the cells in the left half of $D_{10}$
but not shared by the right hand gamete.
With enough mutations, each
round of mitotic germline genome duplication within a single multicellular organism could
in principle be distinguished.

>>>>

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress, however, has led to a
% Controversial, perhaps, but true?
substantial gap opening between theory and practice.
Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
based on exhaustively detailing the ``events'' that occur
in the history of a sample,
are insufficient to describe the outputs of current methods.
Moreover, we argue that the underlying assumption
that all events can be known and precisely estimated
is fundamentally unsuited to the realities
of modern, population-scale datasets.
We propose an alternative mathematical formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample of genomes.
\end{abstract}

@@

% What's the problem?
A major problem currently facing the field is that
classical mathematical formalisms and terminology
cannot adequately describe these vast inferred ARGs.
Fundamentally these formalisms assume that an
ARG is known in complete detail: they
are not suited to describing partial or approximate knowledge.
As we are actively inferring ARGs at the population scale,
and such ARGs can never be fully known,
there is currently a substantial gap between classical representations
and practical application.
The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% This is a digression from the main point of the paragraph, so
% making it a parenthetical aside.
(Note that it is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.)
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

@@

% Moving on from ARG-unary to the more general case
More generally, locally unary nodes,
which can have one or more children in the graph,
are a common and important feature
of many different types of ARG.
As discussed in the previous section,
without these nodes marking the passage
of ancestral material through specific ancestors, the local trees
lack information about events other than local coalescence.
% Not sure if switching example will confuse people here, but would
% like to contrast with example where we don't use unary nodes to
% emphasise the point that we lose info
For example, the local trees for the classical event ARG
depicted in Fig.~\ref{fig-event-arg}B follow the usual conventions
and do not include any information about the recombination
that occurred at node \noderef{d}. Given these two local trees
in isolation,
% (without further information about the corresponding ARG)
we lack specific information about the recombination.
Explicitly recording that node \noderef{d} lies on the
branch joining \noderef{b} to \noderef{e} in the left hand
tree, and \noderef{b} to \noderef{f} in the right hand tree
resolves all ambiguity, and makes the collection of local
trees exactly equivalent to the corresponding ARG (see previous section).
Unary nodes are a vital link between ARGs and local trees, and we
cannot fully reason about how a local tree is embedded in an ARG
without them. As we see in the next two sections, both ARG and locally
unary nodes occur in various scenarios, and are
produced by a range of current inference methods.

@@

% Here we describe recombination features
Considering Fig.~\ref{fig-inferred-args}, we can see that
there is substantial variation in the number of recombination breakpoints
inferred by different methods, with e.g.~\argweaver\ suggesting
far more than the seven required for this dataset under minimal parsimony
assumptions~\citep{song2003parsimonious}.
A sense of the amount of recombination in each ARG is provided by
the node colouring scheme, which shows the number of parents for each node.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event
corresponds to a node with exactly two parents and one child.
As these methods explicitly infer a recombination event for each
breakpoint, the number of breakpoints equals the number of two-parent
(brown) nodes.
In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
straightforward
relationship between the number of nodes with multiple parents
and number of breakpoints along the genome.
In both ARGs the number of breakpoints is smaller
than the number of multiple-parent ARG nodes, showing
that several multiple-parent nodes must share breakpoint positions.
There are also ARG nodes with multiple parents and multiple children,
where one or more recombinations have been pushed down onto a
more recent node.
As a consequence, it may be difficult to condense each transition between trees
in these ARGs into a set of SPR operations.

@@

\section{Discussion}
\label{sec-discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in inference methods, their evaluation, and their application to biological questions.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix~\ref{sec-arg-history}, the term
``ancestral recombination graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition
within the context of stochastic
processes, and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful; we have
tried to clarify and systematise it here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG) encoding, which is more limited in what can be described.
Importantly, gARGs do not require fully precise estimates of
ancestral recombination events,
and allow us to directly express important forms of temporal uncertainty.

Fully decoupling the general concept of an ARG from the coalescent
with recombination (henceforth, ``coalescent'')
is an important step.
While the coalescent has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
One key assumption is that
sample size $n$ is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,
karczewski2020mutational,tanjo2021practical,
halldorsson2022sequences},
and so sample size is an order of magnitude \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from this
assumption, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
A model assuming a single $N_e$ would be a
drastic over-simplification of course, but
even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and myriad
interconnections of migration, were somehow estimated and provided as input,
ARGs sampled from the coalescent cannot capture the complexities
of family structure in these
datasets~\citep[e.g.][]{turnbull2018hundred,Ros-Freixedes2020}.
Another core assumption of the coalescent model is that the genome
(or at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times.
High-quality whole genome assemblies are now available
for many species % Can we say something more concrete here?
with projects under way to obtain
them for tens of thousands more~\citep{darwin2022sequence,lewin2022earth},
and so we can expect inferred ARGs to routinely span
large fractions of a chromosome.

Recent large-scale methods have simplified the inference problem by
making a single, deterministic best-guess at ARG
inference~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes, the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast.
Thus, although it is clearly an oversimplification to arbitrarily
choose \emph{one} best guess, if sample sizes are large
it is not clear that generating many guesses will achieve much.
At the scale of millions of samples,
we could only ever explore the tiniest corner of the incomprehensibly
large space of plausible ARGs.
Therefore, it is important to systematically
describe and use uncertainty about ARG inference, and to
incorporate uncertainty encountered during inference into the returned ARG.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have more than two children (polytomies
representing uncertainty over the ordering of coalescence events, Appendix~\ref{sec-cell-lineages-and-args}) or
more than two parents (representing uncertainty over the ordering
of multiple recombination events, Section~\ref{sec-ARG-simplification}). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.
How this uncertainty can be used in downstream applications
is an open question.

@@

The schematic illustrates an important point about the biological reality of polytomies.
Three lineages coalesce in the left-hand genome of individual $D_{10}$, but do
so as the result of two successive bifurcations. This is \emph{necessarily} so,
because the only known method of reproducing DNA is by (semi-conservative) duplication.
Whether this polytomy is resolvable depends on the available mutational data.
Mutations can occur along any cell lineages. For example,
a mutation in the first cell division of $D_{10}$ could be shared between
the two gametes produced by the cells in the left half of $D_{10}$
but not shared by the right hand gamete.
With enough mutations, each
round of mitotic germline genome duplication within a single multicellular organism could
in principle be distinguished.
